---
title: "Introduction to dplyr"
author: "Paul M. Magwene"
output: beamer_presentation
fontsize: "9pt"    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, comment = "", collapse=TRUE)
options(width=60)
```

```{r}
library(tidyverse)
```


## What is `dplyr`?

`dplyr` is a package that provides a "grammar for data manipulation"

\bigskip

Key "verbs" in the `dplyr` package:

  * `select()`
  * `filter()`
  * `mutate()`
  * `arrange()`
  * `summarize()`
  * `group_by()`
  
\bigskip

All these functions return new data frames instead of modifying existing data frames




## `select()` subsets columns

```{r, echo=TRUE}
names(iris)
```

\medskip

```{r, echo = TRUE}
# select two columns
df <- select(iris, Sepal.Length, Petal.Length)
head(df, 3)
```

\medskip

```{r, echo = TRUE}
# select everything BUT the species column
df <- select(iris, -Species)
head(df, 3)
```

## `select()` has some specialized functions for powerful filtering

```{r, echo = TRUE}
df <- select(iris, starts_with("Petal")) 
head(df, 3)
```

\medskip

```{r, echo = TRUE}
df <- select(iris, ends_with("Length")) 
head(df, 3)
```



## `filter()` selects rows that match criteria

```{r, echo = TRUE}
# get only the I. setosa specimens
df <- filter(iris, Species == "setosa") 
head(df, 3)
```

\medskip

```{r, echo=TRUE}
# filter on mulitple criteria 
df <- filter(iris, Species == "setosa", Sepal.Length < 5)
head(df, 3)
```


## `mutate()` adds or transforms columns

```{r, echo=TRUE}
df <- mutate(iris, Species = str_to_upper(Species))
head(df, 3)
```

## `arrange()` sorts rows according to values of one or more columns

```{r, echo=TRUE}
# sort by Sepal.Length
df <- arrange(iris, Sepal.Length)
head(df, 3)
```

\medskip

```{r, echo=TRUE}
# sort by Sepal.Length then by Petal.Length
df <- arrange(iris, Sepal.Length, Petal.Length)
head(df, 3)
```

## `summarize()` transforms and collapses

`summarize()` applies functions to one or more variables (columns) in the data frame, reducing a vector of values to a single value and returning the results in a data frame

\bigskip

```{r, echo=TRUE}
mean.Lengths <- 
  summarize(iris, 
            avg.Sepal.Length = mean(Sepal.Length), 
            avg.Petal.Length = mean(Petal.Length))

mean.Lengths
```

## `group_by()` is used for conditioning (faceting) and transforming

```{r, echo=TRUE}
# apply grouping
grouped.df <- group_by(iris, Species)

# summarize grouped data frame
mean.by.group <- 
  summarize(grouped.df,
            avg.Sepal.Length = mean(Sepal.Length), 
            avg.Petal.Length = mean(Petal.Length))          
          
mean.by.group
```


## The pipe operator, `%>%`

`dplyr` also provides a new operator called a pipe

  - The pipe operator is `%>%`

Using pipes:

  - `x %>% f()` is equivalent to `f(x)`
  - `x %>% f(y)` is equivalent to `f(x,y)`.

Examples:

  - Single argument function:
    ```{r, echo=TRUE}
pi %>% cos()
```
  
  - For single argument functions, after the pipe you can drop the parentheses:
    ```{r, echo=TRUE}
pi %>% cos  # same as above
```

  - Multi-argument functions
    ```{r, echo=TRUE}
100 %>% log(base=10) # 100 is treated as the first argument
```


## Building pipelines with the pipe operator

The pipe operator allows us to build analysis "pipelines".

\bigskip

A pipeline series of function calls that filter and/or transform our data

\medskip

```{r, echo=TRUE}
letters %>%         # start with letters vector
  str_to_upper %>%  # convert to upper case
  tail(10) %>%      # get last 10 elements
  str_flatten("-")  # join into single string, separated by '-'
```

\bigskip

The pipe operator helps to make our intent clearer, as compared to nested function calls:

\medskip

```{r, echo=TRUE}
str_flatten(tail(str_to_upper(letters), 10), "-")
```

## The `dplyr` verbs functions are designed to work well with piping!


```{r, echo=TRUE}
filtered.Sepal.means <-
  iris %>%
  filter(Species != "virginica") %>%
  group_by(Species) %>%
  summarize(avg.Sepal.Length = mean(Sepal.Length),
            avg.Sepal.Width = mean(Sepal.Width))   


filtered.Sepal.means
```

